GMSE: an R package for generalised management strategy evaluation

### 2018-04-12 notebook/ms/ms.Rmd
Minor edits based on reviewer comments

Many global natural resources, including the biodiversity on which critical ecosystem services depend, are in a state of severe decline [@Dirzo2014; @Hautier2015; @Ceballos2017a; @OConnell2017]. Conservation of biodiversity can be complicated by the immediate need to use natural resources and land area for human livelihood (e.g., food production), causing real or perceived conflicts between biodiversity conservation and food security. This creates a challenging situation for the management of many natural resources [@Redpath2015a]. Given increasing human population size [@Crist2017], the number and intensity of such conflicts are likely to increase into the twenty first century. Effective management tools are therefore needed for the long-term sustainable use of natural resources under the rising demand for food production [@Fischer2017].

To effectively manage natural resources, an adaptive approach allows managers to iteratively update their models and respond flexibly to changing conditions [@Keith2011]. This approach is especially effective when considering all aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of stakeholders [@Bunnefeld2011; @Bunnefeld2014]. Management strategy evaluation (MSE) is a modelling framework, first developed in fisheries, for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016]. Nevertheless, MSE models developed hitherto have been limited in their ability to model human decision-making [@Fulton2011; @Dichmont2017]; manager decisions are typically based on fixed rules, and user behaviour likewise remains fixed over time instead of dynamically responding to changing resource availability and management decisions [@Schluter2012; @Melbourne-Thomas2017]. Here we introduce generalised management strategy evaluation (GMSE), which incorporates a game-theoretic perspective to model the goal-oriented, dynamic decision-making processes of stakeholders.

@@

```{r, echo = FALSE, fig.height = 4, fig.width = 4, fig.align = 'left', fig.cap = "Description of one time step of the generalised management strategy evaluation framework, which is comprised of four separate sub-models."}
mbox <- function(x0, x1, y0, y1){
    xx <- seq(from=x0, to=x1, length.out = 100);
    yy <- seq(from=y0, to=y1, length.out = 100);
    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
    return(list(x=xd, y=yd));
}
par(mar=c(0,0,0,0));
plot(x = 0, y = 0, type = "n", xlim = c(0, 100), ylim = c(0, 100), 
     xaxt = "n", yaxt = "n", xlab = "", ylab = "");
# Manager model 
manger_model_box <- mbox(x0 = 1, x1 = 30, y0 = 70, y1 = 99);
polygon(x = manger_model_box$x, y = manger_model_box$y, lwd = 3);
text(x = 15, y = 90, cex = 1.0, labels = "Manager");
text(x = 15, y = 83, cex = 1.0, labels = "model");
text(x = 5, y = 73, labels="(3)");
# User model
user_model_box <- mbox(x0 = 70, x1 = 99, y0 = 70, y1 = 99);
polygon(x = user_model_box$x, y = user_model_box$y, lwd = 3);
text(x = 85, y = 90, cex = 1.0, labels = "User");
text(x = 85, y = 83, cex = 1.0, labels = "model");
text(x = 74, y = 73, labels="(4)");
# Observation model
observe_model_box <- mbox(x0 = 1, x1 = 30, y0 = 1, y1 = 30);
polygon(x = observe_model_box$x, y = observe_model_box$y, lwd = 3);
text(x = 15, y = 23, cex = 1.0, labels = "Observation");
text(x = 15, y = 15, cex = 1.0, labels = "model");
text(x = 5, y = 4, labels="(2)");
# Resource model
resource_model_box <- mbox(x0 = 70, x1 = 99, y0 = 1, y1 = 30);
polygon(x = resource_model_box$x, y = resource_model_box$y, lwd = 3);
text(x = 85, y = 23, cex = 1.0, labels = "Natural");
text(x = 85, y = 15, cex = 1.0, labels = "resources");
text(x = 85, y = 7,  cex = 1.0, labels = "model");
text(x = 74, y = 4, labels="(1)");
# Game-theoretic (genetic algorithm) model
game_model_box <- mbox(x0 = 36, x1 = 65, y0 = 36, y1 = 65);
polygon(x = game_model_box$x, y = game_model_box$y, lwd = 3);
text(x=50, y=58, labels="Genetic", cex=1.0);
text(x=50, y=51, labels="algorithm", cex=1.0);
sc1 <- seq(from = 0, to = pi, by = 0.001);
sc2 <- seq(from = pi, to = 2*pi, by = 0.001);
points(x = 2*cos(sc1) + 41, y = 42 + 2*sin(sc1), lwd = 1.5, type = "l");
points(x = 2*cos(sc2) + 41, y = 40 + 2*sin(sc2), lwd = 1.5, type = "l");
arrows(x0 = 43, x1 = 43, y0 = 42, y1 = 41, lwd = 1.5, length = 0.06);
arrows(x0 = 39, x1 = 39, y0 = 40, y1 = 41, lwd = 1.5, length = 0.06);
text(x=43, y=42, labels="Multiple", cex=0.7, pos = 4);
text(x=43, y=39, labels="Generations", cex=0.7, pos = 4);
# Arrows indicating direction of GMSE model
arrows(x0=15, x1=15, y0=30, y1=70, lwd=2, length=0.15);
arrows(x0=30, x1=70, y0=85, y1=85, lwd=2, length=0.15);
arrows(x0=85, x1=85, y0=70, y1=30, lwd=2, length=0.15);
arrows(x0=70, x1=30, y0=15, y1=15, lwd=2, length=0.15);
arrows(x0=30, x1=36, y0=70, y1=65, lwd=2, length=0.075, code=3);
arrows(x0=65, x1=70, y0=65, y1=70, lwd=2, length=0.075, code=3);
text(x=50, y=90, labels="Policy", cex = 1.0);
text(x=50, y=20, labels="Monitoring", cex = 1.0);
text(x=10, y=48, labels="Indicators", srt=90, cex = 1.0);
text(x=90, y=52, labels="Actions", srt=-90, cex = 1.0);
```

@@

In each new call of the genetic algorithm, a unique population of managers or users with random strategies is temporarily initialised. In each generation of the genetic algorithm, these strategies crossover and mutate; when this results in strategies that are over-budget, expenditures are iteratively decreased at random until budget constraints are satisfied. A fitness function then evaluates each strategy in the population, and a tournament is used to select the next generation of strategies [@Hamblin2013]. The genetic algorithm terminates when a minimum number of generations has passed and the increase in the fitness of the fittest strategy between the current and previous generation is below some threshold. The highest fitness strategy in the population then becomes the stakeholder's new strategy. Overall, a single call of the genetic algorithm thereby simulates the process of thinking and decision-making for one manager or user.

@@

```{r, echo = FALSE, fig.height = 6, fig.align = 'left', fig.cap = "Results of an example simulation illustrating the management of a protected resource that exploits the land of five farmers. The upper left panel shows locations of resources (black dots) on the landscape in the final time step of the simulation. The upper right panel shows the same landscape broken down into five differently coloured regions, which correspond to areas of land owned by each of the five farmers. The middle left panel shows the actual abundance of resources (black solid line), and the abundance of resources as estimated by the manager (blue solid line; shading indicates 95 percent confidence intervals), over time. The horizontal dotted red and blue lines show the resource carrying capacity enacted on adult mortality and the manager's target for resource abundance, respectively. The orange line shows the total percent yield of landscape cells. The middle right panel shows total percent yield of landscape cells for each individual farmer, differentiated by colour, where line colours correspond to areas of the landscape in the upper right panel. The lower left panel shows the cost of farmers performing actions over time, as set by the manager. The lower right panel shows the total number of actions attempted to be performed by all farmers over time (some actions might be unsuccessful if resources are unavailable on a farmer's land to cull or scare, so, e.g., culling actions might be larger than resources actually culled)."}
plot_gmse_results_vign(res = sim$resource, obs = sim$observation, 
                       land = sim$land, agents = sim$agents, paras = sim$paras,
                       ACTION = sim$action, COST = sim$cost);
```

>>>>

Many global natural resources, including the biodiversity on which critical ecosystem services depend, are in a state of severe decline [@Dirzo2014; @Hautier2015; @Ceballos2017a; @OConnell2017]. Conservation of biodiversity can be complicated by the immediate need to use natural resources and land area for human livelihood, causing real or perceived conflicts between biodiversity conservation and agricultural production. This creates a challenging situation for the management of many natural resources [@Redpath2015a]. Given increasing human population size [@Crist2017], the number and intensity of such conflicts are likely to increase into the twenty first century. Effective management tools are therefore needed for the long-term sustainable use of natural resources under the rising demand for food production [@Fischer2017].

To effectively manage natural resources, an adaptive approach allows managers to iteratively update their models and respond flexibly to changing conditions [@Keith2011]. This approach is especially effective when considering multiple aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of stakeholders [@Bunnefeld2011; @Bunnefeld2014]. Management strategy evaluation (MSE) is a modelling framework, first developed in fisheries, for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016]. Nevertheless, MSE models developed hitherto have been limited in their ability to model human decision-making [@Fulton2011; @Dichmont2017]; manager decisions are typically based on fixed rules, and user behaviour likewise remains fixed over time instead of dynamically responding to changing resource availability and management decisions [@Schluter2012; @Melbourne-Thomas2017]. Here we introduce generalised management strategy evaluation (GMSE), which incorporates a game-theoretic perspective to model the goal-oriented, dynamic decision-making processes of stakeholders.

@@

```{r, echo = FALSE, fig.height = 4, fig.width = 4, fig.align = 'left', fig.cap = "Description of one time step of the generalised management strategy evaluation framework, which is comprised of four separate sub-models."}
mbox <- function(x0, x1, y0, y1){
    xx <- seq(from=x0, to=x1, length.out = 100);
    yy <- seq(from=y0, to=y1, length.out = 100);
    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
    return(list(x=xd, y=yd));
}
par(mar=c(0,0,0,0));
plot(x = 0, y = 0, type = "n", xlim = c(0, 100), ylim = c(0, 100), 
     xaxt = "n", yaxt = "n", xlab = "", ylab = "");
# Manager model 
manger_model_box <- mbox(x0 = 1, x1 = 30, y0 = 70, y1 = 99);
polygon(x = manger_model_box$x, y = manger_model_box$y, lwd = 3);
text(x = 15, y = 90, cex = 1.0, labels = "Manager");
text(x = 15, y = 83, cex = 1.0, labels = "model");
text(x = 5, y = 73, labels="(3)");
# User model
user_model_box <- mbox(x0 = 70, x1 = 99, y0 = 70, y1 = 99);
polygon(x = user_model_box$x, y = user_model_box$y, lwd = 3);
text(x = 85, y = 90, cex = 1.0, labels = "User");
text(x = 85, y = 83, cex = 1.0, labels = "model");
text(x = 74, y = 73, labels="(4)");
# Observation model
observe_model_box <- mbox(x0 = 1, x1 = 30, y0 = 1, y1 = 30);
polygon(x = observe_model_box$x, y = observe_model_box$y, lwd = 3);
text(x = 15, y = 23, cex = 1.0, labels = "Observation");
text(x = 15, y = 15, cex = 1.0, labels = "model");
text(x = 5, y = 4, labels="(2)");
# Resource model
resource_model_box <- mbox(x0 = 70, x1 = 99, y0 = 1, y1 = 30);
polygon(x = resource_model_box$x, y = resource_model_box$y, lwd = 3);
text(x = 85, y = 23, cex = 1.0, labels = "Natural");
text(x = 85, y = 15, cex = 1.0, labels = "resources");
text(x = 85, y = 7,  cex = 1.0, labels = "model");
text(x = 74, y = 4, labels="(1)");
# Game-theoretic (genetic algorithm) model
game_model_box <- mbox(x0 = 36, x1 = 65, y0 = 36, y1 = 65);
polygon(x = game_model_box$x, y = game_model_box$y, lwd = 3);
text(x=50, y=58, labels="Genetic", cex=1.0);
text(x=50, y=51, labels="algorithm", cex=1.0);
sc1 <- seq(from = 0, to = pi, by = 0.001);
sc2 <- seq(from = pi, to = 2*pi, by = 0.001);
points(x = 2*cos(sc1) + 41, y = 42 + 2*sin(sc1), lwd = 1.5, type = "l");
points(x = 2*cos(sc2) + 41, y = 40 + 2*sin(sc2), lwd = 1.5, type = "l");
arrows(x0 = 43, x1 = 43, y0 = 42, y1 = 41, lwd = 1.5, length = 0.06);
arrows(x0 = 39, x1 = 39, y0 = 40, y1 = 41, lwd = 1.5, length = 0.06);
text(x=43, y=42, labels="Multiple", cex=0.7, pos = 4);
text(x=43, y=39, labels="Iterations", cex=0.7, pos = 4);
# Arrows indicating direction of GMSE model
arrows(x0=15, x1=15, y0=30, y1=70, lwd=2, length=0.15);
arrows(x0=30, x1=70, y0=85, y1=85, lwd=2, length=0.15);
arrows(x0=85, x1=85, y0=70, y1=30, lwd=2, length=0.15);
arrows(x0=70, x1=30, y0=15, y1=15, lwd=2, length=0.15);
arrows(x0=30, x1=36, y0=70, y1=65, lwd=2, length=0.075, code=3);
arrows(x0=65, x1=70, y0=65, y1=70, lwd=2, length=0.075, code=3);
text(x=50, y=90, labels="Policy", cex = 1.0);
text(x=50, y=20, labels="Monitoring", cex = 1.0);
text(x=10, y=48, labels="Indicators", srt=90, cex = 1.0);
text(x=90, y=52, labels="Actions", srt=-90, cex = 1.0);
```

@@

Overall, a single call of the genetic algorithm simulates the process of thinking and decision-making for one manager or user. In each new call of the genetic algorithm, a unique population of temporary manager or user strategies is initialised. In each iteration of the genetic algorithm, these strategies crossover and mutate; when this results in strategies that are over-budget, expenditures are iteratively decreased at random until budget constraints are satisfied. A fitness function then evaluates each strategy in the population, and a tournament is used to select the next iteration of strategies [@Hamblin2013]. The genetic algorithm terminates when a minimum number of iterations has passed and the increase in the fitness of the fittest strategy between the current and previous iteration is below some threshold. The highest fitness strategy in the population then becomes the stakeholder's new strategy.

@@

```{r, echo = FALSE, fig.height = 6, fig.align = 'left', fig.cap = "Results of an example simulation illustrating the management of a protected resource that exploits the land of five farmers. The upper left panel shows locations of resources (black dots) on the landscape in the final time step of the simulation (multiple resources can occur on the same landscape cell). The upper right panel shows the same landscape broken down into five differently coloured regions, which correspond to areas of land owned by each of the five farmers. The middle left panel shows the actual abundance of resources (black solid line), and the abundance of resources as estimated by the manager (blue solid line; shading indicates 95 percent confidence intervals), over time. The horizontal dotted red and blue lines show the landscape-level resource carrying capacity enacted on adult mortality and the manager's target for resource abundance, respectively. The orange line shows the total percent yield of landscape cells. The middle right panel shows total percent yield of landscape cells for each individual farmer, differentiated by colour, where line colours correspond to areas of the landscape in the upper right panel. The lower left panel shows the cost of farmers performing actions over time, as set by the manager. The lower right panel shows the total number of actions attempted to be performed by all farmers over time (some actions might be unsuccessful if resources are unavailable on a farmer's land to cull or scare, so, e.g., culling actions might be larger than resources actually culled)."}
plot_gmse_results_vign(res = sim$resource, obs = sim$observation, 
                       land = sim$land, agents = sim$agents, paras = sim$paras,
                       ACTION = sim$action, COST = sim$cost);
```
